---
title: "Visual analysis"
author: "Jei Pratheesh"
date: "2025-04-30"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(scales)
library(ggplot2)
library(janitor)
```

# Load Dataset

```{r}
data <- read_csv("../data/Student_Mental_Stress_and_Coping_Mechanisms.csv")

# Clean column names for ease of use
data <- clean_names(data)
```

```{r}
# Check for missing values
colSums(is.na(data))

# Summary of numeric variables
summary(select(data, where(is.numeric)))

# Check for implausible values
data %>%
  filter(academic_performance_gpa > 4 |
         study_hours_per_week > 100 |
         social_media_usage_hours_per_day > 24 |
         sleep_duration_hours_per_night > 24 |
         physical_exercise_hours_per_week > 100)
```


# Feature Engineering

## 1. Weekly Aggregates

```{r}
data <- data %>%
  mutate(
    weekly_social_media_hours = social_media_usage_hours_per_day * 7,
    weekly_sleep_hours = sleep_duration_hours_per_night * 7,
    weekly_exercise_hours = physical_exercise_hours_per_week
  )
```

## 2. Academic Load Index (Raw and Normalised)

```{r}
data <- data %>%
  mutate(
    academic_load_index = academic_performance_gpa * study_hours_per_week,
    academic_load_index_scaled = rescale(academic_load_index, to = c(0, 1))
  )
```

## 3. Stress Category (Based on Mental Stress Level)

```{r}
data <- data %>%
  mutate(
    stress_category = case_when(
      mental_stress_level <= 3 ~ "Low",
      mental_stress_level <= 6 ~ "Moderate",
      mental_stress_level >= 7 ~ "High",
      TRUE ~ NA_character_
    )
  )
```

## 4. Healthy Lifestyle Score

```{r}
data <- data %>%
  mutate(
    healthy_score = as.integer(sleep_duration_hours_per_night >= 7) +
                    as.integer(physical_exercise_hours_per_week >= 3) +
                    as.integer(diet_quality >= 3)
  )
```

## 5. Categorical Coping Type

```{r}
data <- data %>%
  mutate(coping_type = case_when(
    str_detect(stress_coping_mechanisms, regex("meditation|breathing|yoga|alone", ignore_case = TRUE)) ~ "Emotional",
    str_detect(stress_coping_mechanisms, regex("talking|family|friends", ignore_case = TRUE)) ~ "Social",
    str_detect(stress_coping_mechanisms, regex("reading|drawing|writing", ignore_case = TRUE)) ~ "Creative",
    str_detect(stress_coping_mechanisms, regex("music|watching sports", ignore_case = TRUE)) ~ "Relaxation",
    str_detect(stress_coping_mechanisms, regex("travelling|nature", ignore_case = TRUE)) ~ "Adventure",
    str_detect(stress_coping_mechanisms, regex("social media", ignore_case = TRUE)) ~ "Distraction",
    str_detect(stress_coping_mechanisms, regex("exercise", ignore_case = TRUE)) ~ "Physical",
    TRUE ~ "Other"
  ))
```

## 6. At-Risk Flag

```{r}
data <- data %>%
  mutate(
    at_risk = ifelse(
      mental_stress_level >= 7 &
      sleep_duration_hours_per_night <= 5 &
      diet_quality <= 2, 1, 0)
  )
```

## Feature Preview

```{r}
glimpse(data)
summary(data)
```

# Visual analysis

## Boxplot of Weekly Sleep Hours by Stress Category

```{r}
ggplot(data, aes(x = stress_category, y = sleep_duration_hours_per_night, fill = stress_category)) +
  geom_boxplot(outlier.color = "red", alpha = 0.7) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Daily Sleep Duration Across Stress Categories",
    x = "Stress Category",
    y = "Sleep Duration (Hours per Night)",
    fill = "Stress Category"
  ) +
  theme_minimal()
```

## GPA by Stress Category

```{r}
ggplot(data, aes(x = stress_category, y = academic_load_index_scaled, fill = stress_category)) +
  geom_boxplot(outlier.color = "red", alpha = 0.7) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Academic Load Index (Scaled) by Stress Category",
    x = "Stress Category",
    y = "Academic Load Index (0–1)",
    fill = "Stress Category"
  ) +
  theme_minimal()
```

## Coping Mechanism Type by Stress Category

```{r}
ggplot(data, aes(x = stress_category, fill = coping_type)) +
  geom_bar(position = "fill") +
  scale_fill_brewer(palette = "Paired") +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Coping Mechanism Proportions by Stress Category",
    x = "Stress Category",
    y = "Proportion of Students",
    fill = "Coping Type"
  ) +
  theme_minimal()
```

## Healthy Score by Stress Category

```{r}
risk_counts <- data %>%
  mutate(risk_status = ifelse(at_risk == 1, "At-Risk", "Not At-Risk")) %>%
  count(risk_status) %>%
  mutate(
    percent = round(n / sum(n) * 100, 1),
    label = paste0(risk_status, ": ", percent, "%")
  )

# Final plot: At-Risk vs Not At-Risk students (horizontal bar)
ggplot(risk_counts, aes(x = reorder(risk_status, n), y = n, fill = risk_status)) +
  geom_col(width = 0.6) +
  geom_text(aes(
    label = label,
    hjust = ifelse(risk_status == "At-Risk", -0.2, 1.1)
  ),
  vjust = 0.5,
  size = 5,
  color = "black") +
  scale_fill_manual(values = c("At-Risk" = "#66c2a5", "Not At-Risk" = "#fc8d62")) +
  labs(
    title = "Distribution of At-Risk vs Not At-Risk Students",
    x = NULL,
    y = "Number of Students",
    fill = "Student Group"
  ) +
  coord_flip() +
  expand_limits(y = max(risk_counts$n) * 1.2) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )
```

## Social Media Usage Across Stress Levels

```{r}
ggplot(data, aes(x = social_media_usage_hours_per_day)) +
  geom_histogram(binwidth = 1, fill = "#80b1d3", color = "white") +
  facet_wrap(~ stress_category) +
  labs(
    title = "Social Media Usage by Stress Category",
    x = "Hours per Day on Social Media",
    y = "Number of Students"
  ) +
  theme_minimal()
```

```{r}
ggplot(data, aes(x = social_media_usage_hours_per_day, fill = stress_category)) +
  geom_histogram(binwidth = 1, color = "white", alpha = 0.8) +
  facet_wrap(~ stress_category, ncol = 1, strip.position = "top") +
  scale_fill_brewer(palette = "Set2") +
  scale_x_continuous(breaks = seq(0, 10, 1)) +
  labs(
    title = "Daily Social Media Usage Across Stress Categories",
    x = "Hours per Day on Social Media",
    y = "Number of Students",
    fill = "Stress Category"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 12, face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "none"
  )
```

```{r}
# Create social media usage buckets
data <- data %>%
  mutate(sm_usage_group = case_when(
    social_media_usage_hours_per_day <= 2 ~ "0–2 hrs",
    social_media_usage_hours_per_day <= 5 ~ "3–5 hrs",
    TRUE ~ "6+ hrs"
  ))

# Plot grouped bar chart
ggplot(data, aes(x = sm_usage_group, fill = stress_category)) +
  geom_bar(position = "dodge") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Stress Category Distribution by Social Media Usage Group",
    x = "Daily Social Media Usage",
    y = "Number of Students",
    fill = "Stress Category"
  ) +
  theme_minimal()
```

## Coping Type Proportions (Overall)

```{r}
# Count and calculate percentages
coping_counts <- data %>%
  count(coping_type) %>%
  mutate(
    percent = round(n / sum(n) * 100, 1),
    label = paste0(coping_type, "\n", percent, "%")
  )

# Donut chart
ggplot(coping_counts, aes(x = 2, y = n, fill = coping_type)) +
  geom_col(width = 1, color = "white") +
  coord_polar(theta = "y") +
  geom_text(aes(label = label), position = position_stack(vjust = 0.5), size = 4) +
  xlim(0.5, 2.5) +  # Creates donut hole
  scale_fill_brewer(palette = "Paired") +
  labs(
    title = "Overall Coping Strategy Usage",
    fill = "Coping Type"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )
```

